Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-07-12 concurrency/Verification of Concurrent Binary Search Tree.tex
doing some editing

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}

@@

\title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
\author{Roshan Sharma}
\author{Shengyi Wang}
\author{Lennart Beringer}
\author{William Mansky}

@@

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

@@

technical material: definition of sync\_inv, sync\_commit and sync\_rollback

@@

\subsubsection{Ghost States and Range}
 The main challenge in proving the properties of concurrent data structure is to abstract the data structure in a way that helps to reason about it in the concurrent setting. Krishna et al. \cite{krishna2017flow} mention in their paper that we can abstract binary search tree using appropriate bounds in each node: we can use the product of \emph{lower-bound} and \emph{upper-bound} on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree. The \emph{lower-bound} and \emph{upper-bound} help us to prove that we perform the operation in the right node: for instance, if we $\mathsf{insert}$ a key in the tree, then key should be inside those two bounds. We use the term \emph{range} to represent the product of lower- and upper-bound. \ref{range_bst} shows an example of BST with each node labeled with appropriate bound propagated all the way from \emph{root} to empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds. For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound). This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state. For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node. Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$. During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40). We introduce the $range$ in our proof using the ghost variables. This $ranges$ along with other information about the node represent the abstract state of the data structure. We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
\textbf{Per-node Ghost State}: The per-node ghost state is the product of the range and other information about node (key, value, and ghost name for its child nodes), created for each node. In VST, to use any Coq type as a ghost state, we need to create an instance of \texttt{Ghost} typeclass, \emph{separation algebra} with associated \texttt{join} relation, with an additional \texttt{valid} predicate marking those elements of the algrebra that can be used in ghost assertions. \texttt{Ghost} typeclass is related to the PCMs described in \ref{chapter2}. For instance, ghost instance for range can be created as follows:
\begin{verbatim}
 Program Instance range_ghost : Ghost :=
  { G := (number*number); valid g := True; Join_G a b c := c =  merge_range a b }.
\end{verbatim}
Where \texttt{number} $\in (-\infty,+\infty)$ and $\texttt{merge\_range}$ is a function that merges two ranges into one. We extend this range information with the other information about the node: key, value and $\texttt{gname}$ for its child. We define our total ghost state as follows:
\begin{verbatim}
Definition ghost_info : Type := (key * val * gname * gname)%type.
Instance node_ghost : Ghost := prod_PCM range_ghost (exclusive_PCM (option ghost_info)).
\end{verbatim}
The \texttt{exclusive\_PCM} is an exlusive algebra, which means that \texttt{Some} only joins with \texttt{None}, not with another \texttt{Some} (even the same one). Now, we use ghost assertion to represent this instance. This ghost state follow \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a "reference" copy that records all of the contributions. VST provides \texttt{my\_half} and \texttt{public\_half} assertions for this: contribution element $\mathsf{my\_half}\ \mathit{g}\ \mathit{sh}\ a$ and reference element $\mathsf{public\_half}\ \mathit{g}\ r$. To join two elements, we combine the shares and values of the contributions, and require that the elements contain at most one reference between them to ensure the uniqueness of the reference value. When \texttt{my\_half} element has the full share, it is guaranteed to be equal to the reference element. In our case, \texttt{my\_half} assertion will be local to a thread (that means it will stay in private pre- and postcondition of atomic specification), and  \texttt{public\_half} will be shared with all threads (that means public pre-and postcondition).

\textbf{Overall Ghost State}: We have constructed the per-node ghost state for each node, using unique ghost variable for each node. But if we are given a ghost variable (\emph{gname}) to access the resources in the node labeled with that variable, we do not any clue about whether or not that variable exists in the tree. So, we need another kind of ghost state, which represents the a set of all per-node ghost variable. Again We can use ``reference" pattern to define this ghost state:
\begin{verbatim}
Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
\end{verbatim}
We use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part of the atomic specification respectively. The global invariant, which we introduced in our atomic specification as $\treerep$ predicate above, ties together $\texttt{public\_half}$ and $\texttt{ghost\_ref}$ as follows:
\begin{align*}\treerep(T,g) \triangleq [\ \forall n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively. Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree. We proved the following lemma for this:
$$
Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
$$
\subsubsection{Combining all pieces in atomic specification}
We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and over-all ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

The $\nodeboxrep$ is a predicate describing the local information about the node almost similar to the predicate we defined for the safety proof in Section \ref{safety}. In this case, the lock invariant of recursive lock have, along with the information about the children node,$\mathsf{my\_half}$ predicate which holds the partial information describing the current thread's contribution to the shared state.

@@

The atomic specification for any binary search tree's operation \texttt{op} can be written as follows:
\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
\end{align*}

Where $t$ and $t'$ are the abstract state of a tree before and after the function's execution.

>>>>

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}

@@

\title{Verifying Fine-Grained and Lock-Free Binary Search Trees}
\author{Roshan Sharma}
\author{Shengyi Wang}
\author{Alexander Oey}
\author{Lennart Beringer}
\author{William Mansky}

@@

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

@@

@@

\subsubsection{Ghost States and Range}
 The main challenge in proving the properties of concurrent data structure is to abstract the data structure in a way that helps to reason about it in the concurrent setting. Krishna et al. \cite{krishna2017flow} mention in their paper that we can abstract binary search tree using appropriate bounds in each node: we can use the product of \emph{lower-bound} and \emph{upper-bound} on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree. The \emph{lower-bound} and \emph{upper-bound} help us to prove that we perform the operation in the right node: for instance, if we $\mathsf{insert}$ a key in the tree, then key should be inside those two bounds. We use the term \emph{range} to represent the product of lower- and upper-bound. \ref{range_bst} shows an example of BST with each node labeled with appropriate bound propagated all the way from \emph{root} to empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds. For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound). This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state. For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node. Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$. During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40). We introduce the $range$ in our proof using the ghost variables. This $ranges$ along with other information about the node represent the abstract state of the data structure. We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
\textbf{Per-node Ghost State}: The per-node ghost state is includes a node's range and other information about the node: its current key, value, and ghost names of its child nodes. In VST, we can use a type as ghost state by proving an instance of the \texttt{Ghost} typeclass, analogous to the \emph{resource algebras} of Iris. The essential part is a \emph{join} relation describing how two pieces of ghost state combine. We define the join of two ranges to be the range that contains both of them, and use an \emph{exclusive} algebra for the rest of the data, so that only one thread knows the current state of a node at a time. 
Ghost state is shared between the per-node lock invariants and the abstract state of the data structure using the \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a ``reference'' copy that records all of the contributions. In VST, this is implemented with assertions $\mathsf{my\_half}\ g\ \mathit{sh}\ a$ and $\mathsf{public\_half}\ g\ r$, where $g$ is the node's identifier ($\gnamety$), $a$/$r$ is the value of the ghost state, and $\mathit{sh}$ is a share describing how much of the reference value is known: if $\mathit{sh}$ is the full share, then we know that $a = r$. In the proof, the \texttt{my\_half} assertion for a node will be held by the node's lock, and \texttt{public\_half} will be held by the shared state.

\textbf{Overall Ghost State}: We now have per-node ghost state for each node, indexed by a ghost name $g_n$. But if we are given an arbitrary $\gnamety$ for a node, we do not know whether that $\gnamety$ exists in the tree. We need another kind of ghost state to hold the set of all nodes in the current tree. Again we can use the ``reference'' pattern to define this ghost state, where the abstract state $\mathsf{ghost\_ref}\ s$ holds the full set of nodes, and each node's lock invariant holds an $\mathsf{in\_tree\ g_n}$ assertion guaranteeing that $g_n$ is in $s$.%notation for gnames
\ignore{\begin{verbatim}
Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
\end{verbatim}}
\ignore{The global invariant, which we introduced in our atomic specification as the $\treerep$ predicate above, ties together the pieces as follows:
\begin{align*}\treerep(T,g) \triangleq *_n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively. Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree. We proved the following lemma for this:
$$
Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
$$}
\subsubsection{Handle and Abstract State}
Now we are ready to define the $\treerep$ and $\nodeboxrep$ predicates.

describe lock invariant

We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

@@

The atomic specification for any binary search tree's operation \texttt{op} should then be
\begin{align*}\forall t.\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t,g)\rangle\ \texttt{op}(p)\ \langle \nodeboxrep(p,g,g_{\mathit{root}})\ |\ \treerep(t', g)\rangle
\end{align*}
where $t$ and $t'$ are the abstract state of a tree before and after the function's execution. Now we are prepared to prove that each of our BST functions satisfies its specification.
